---
title: "project1_deseq2"
output:
  pdf_document: default
  html_document: default
---
# imports
```{r set-up}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library(dplyr)
library(fgsea)
library(ggbeeswarm)
library(gplots)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(tidyverse)
library(sessioninfo)
```

# package versions
```{r}
session_info()
```

# filter count data barplot
```{r}
# read in counts
counts <- read_csv("./results/counts.csv")

# filter out genes with < 10 reads
nonzero_genes <- rowSums(counts[-1]) >= 10
filtered_counts <- counts[nonzero_genes,] %>% arrange()
head(filtered_counts)

barplot_no_filter <- counts %>%
    pivot_longer(cols = starts_with("exp") | starts_with("control"), names_to = "condition", values_to = "value") %>%
    pivot_wider(names_from = "gene", values_from = "value") %>%
    mutate(total_gene_count = rowSums(.[,-1])) %>%
    dplyr::select(condition, total_gene_count)

barplot_filter <- filtered_counts %>%
    pivot_longer(cols = starts_with("exp") | starts_with("control"), names_to = "condition", values_to = "value") %>%
    pivot_wider(names_from = "gene", values_from = "value") %>%
    mutate(total_gene_count = rowSums(.[,-1])) %>%
    dplyr::select(condition, total_gene_count)

merged_barplot <- dplyr::right_join(barplot_no_filter, barplot_filter, by = "condition") %>%
    rename("no_filter" = total_gene_count.x, "filter" = total_gene_count.y) %>%
    pivot_longer(!condition, names_to = "filtration", values_to = "count") %>%
    ggplot(aes(fill = filtration, x = condition, y = count)) +
    geom_bar(position="dodge", stat="identity")
# + coord_cartesian(ylim = c(1.6e8, 1.603e8))

merged_barplot

dim(counts)
# 63241 genes
dim(filtered_counts)
# 29444 genes
```

# DESeq2/top 10 significant genes
```{r}
# make count matrix, coldata, design matrix
count_matrix <- filtered_counts[-1] %>% as.matrix()
row.names(count_matrix) <- filtered_counts$gene

head(count_matrix)

coldata <- tibble(
  sample_name=colnames(filtered_counts[-1])) %>%
  separate(sample_name,c("condition","replicate"),sep="_",remove=FALSE) %>%
  mutate(condition = as_factor(condition))

coldata

design <- formula(~ condition)

# run deseq2
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = coldata,
  design = design
)

dds <- DESeq(dds)
res <- results(dds)

# join deseq res to hgnc symbols
hgnc_ids <- read.delim("./results/star/geneInfo.tab", sep = "\t", header = FALSE, skip = 1)
colnames(hgnc_ids) <- c("ensembl_id", "hgnc_symbol", "description")

res_hgnc <- res %>%
    as_tibble(rownames = "ensembl_id") %>% 
    arrange(padj) %>%
    dplyr::left_join(., hgnc_ids, by = "ensembl_id")

# top 10 most significant by padj
top_ten <- res_hgnc %>%
    slice_head(n = 10)

print(top_ten, width = Inf)

top_ten$hgnc_symbol
```

# number significant genes at padj <= 0.05
```{r}
padj_threshold <- 0.05

num_sig_genes <- res %>%
    as.data.frame() %>%
    filter(padj <= padj_threshold)

length(num_sig_genes$padj)
```
# heatmap and pca
```{r}
# transform counts w rlog
transformed_counts <- rlog(dds, blind=FALSE)

# sample distances and heatmap
sample_dists <- dist(t(assay(transformed_counts)))
sampleDistMatrix <- as.matrix(sample_dists)
rownames(sampleDistMatrix) <- transformed_counts$sample_name
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
  clustering_distance_rows=sample_dists,
  clustering_distance_cols=sample_dists,
  col=colors)

# pca and biplot
plotPCA(transformed_counts, intgroup=c("sample_name"))
```
# fgsea and barplot(3F)
```{r}
# join hgnc symbols with deseq2 results
res_log2f_sorted <- res_hgnc %>% 
dplyr::select("hgnc_symbol", "log2FoldChange") %>%
filter(!is.na(hgnc_symbol)) %>%
arrange(desc(log2FoldChange))

# created named vector for ranked genes and read in .gmt file for pathways
ranked_genes <- setNames(res_log2f_sorted$log2FoldChange, res_log2f_sorted$hgnc_symbol)
pathways <- gmtPathways("./refs/c2.cp.reactome.v2024.1.Hs.symbols.gmt")

# fgsea analysis
fgsea_res <- fgsea(pathways, ranked_genes, minSize=15, maxSize=500) %>% 
as_tibble() %>%
arrange(desc(NES))

# filter and plot top 10 down- and up-regulated pathways by NES score
top10 <- head(fgsea_res, 10)
bottom_10 <- tail(fgsea_res, 10)

combined_nes <- bind_rows(top10, bottom_10)

combined_nes$pathway

barplot <- ggplot(combined_nes, aes(x = pathway, y = NES)) + geom_bar(stat = "identity") +
theme(axis.text.y = element_text(size = 5), legend.position = "none") +
coord_flip() 

barplot
```

### volcano plot(3C)
```{r}
library(scales)

volcano_plot <- res_hgnc %>%
  as_tibble() %>%
  arrange(desc(padj)) %>%
  ggplot(mapping = aes(log2FoldChange, -log10(padj))) + 
  geom_point(aes(color = ifelse(abs(log2FoldChange) > 0.2 & padj < 0.05, 
    ifelse(log2FoldChange > 0, "Up", "Down"), "NS"))) +
  geom_label( 
    data = res_hgnc %>% filter(hgnc_symbol %in% c("KRAS", "ELMO1", "PAX6", "LAMB2", "SPP1", "BAX", "GAB2", "NOS3")),
    aes(label=hgnc_symbol)
  ) +
  scale_color_manual(values = c("Up" = "#F8766D", "Down" = "#619CFF", "NS" = "black")) +
  geom_vline(xintercept = -0.25, linetype = 3) +
  geom_vline(xintercept = 0.25, linetype = 3) +
  xlab("Expression difference (log2FC)") +
  ylab("Significance Value (-log10 p value)") +
  ggtitle("S5") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_x_continuous(limits = c(-10, 10), oob = scales::oob_squish) +
  scale_y_continuous(limits = c(0, 15), oob = scales::oob_squish) +
  annotate("text", x = -6, y = 16, label = "DOWN", color = "#619CFF", size = 6) +
  annotate("text", x = 6, y = 16, label = "UP", color = "#F8766D", size = 6)

volcano_plot
```
